---
title: 'Lab 4'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

*Agenda*: Importing and manipulating data; writing functions to estimate
parameters; writing functions to check model fit.

We look at a more systematic data source than Forbes magazine, the World
Top Incomes Database hosted by the Paris School of Economics. This is
derived from income tax reports, and compiles information about the very
highest incomes in various countries over time, trying as hard as
possible to produce numbers that are comparable across time and space.

For most countries in most time periods, the upper end of the income
distribution $roughly$ follows a ***Pareto distribution***, with
probability density function

$$ f(x) = \frac{(a-1)}{x_{min}}\bigg( \frac{x}{x_{min}} \bigg)^{-a}$$

for incomes $x \ge x_{min}$. (Typically, $x_{min}$ is large enough that
only the richest $3-4\%$ of the population falls above it.) As the
***Pareto exponent*** $a$ gets smaller, the distribution of income
becomes more unequal, that is, more of the population's total income is
concentrated among the very richest people.

The proportion of people whose income is at least $x_{min}$ whose income
is $also$ at or above any level $w \ge x_{min}$ is thus

$$Pr(X \ge w) = \int_{w}^{\infty} f(x)dx = \int_{x}^{infty} \frac{(a-1)}{x_{min}}\bigg( \frac{x}{x_{min}} \bigg)^{-a} dx = \bigg( \frac{w}{x_{min}} \bigg)^{-a+1} $$

We will use this to estimate how income inequality changed in the US
over the last hundred years or so. (Whether the trends are good or bad
or a mix is beyond our scope here.)

WTID exports its data sets as .xlsx spreadsheets. For this lab session,
we have extracted the relevant data and saved it as wtid-report.csv.

# Part I

1.  Open the file and make a dataset containing only the year, "P99",
    "P99.5" and "P99.9" variables; these are the income levels which put
    one at the 99th, 99.5th, and 99.9th, percentile of income. What was
    P99 in 1972? P99.5 in 1942? P99.9 in 1922? You must identify these
    using your code rather than looking up the values manually. (You may
    want to modify the column names to make some of them shorter.)

    ```{r}
    #part1
    data <- read.csv("wtid-report.csv")
    dim(data)
    head(data)

    colnames(data)
    new.data <- data[,-c(1,3,4,8)]
    dim(new.data)
    head(new.data)
    colnames(new.data)
    colnames(new.data) <- c("Year","P99","P99.5","P99.9")
    colnames(new.data)

    #part2
    new.data$P99[which(new.data$Year==1972)]
    new.data$P99.5[which(new.data$Year==1942)]
    new.data$P99.9[which(new.data$Year==1922)]

    ```

2.  Provide summary of all three variables and make the corresponding
    plots.

    ```{r}
    summary(new.data$P99)
    plot(new.data$P99)

    summary(new.data$P99.5)
    plot(new.data$P99.5)

    summary(new.data$P99.9)
    plot(new.data$P99.9)
    ```

3.  One can show from the earlier equations that one can estimate the
    exponent by the formula

$$ a = 1 - \frac{log \hspace{0.1cm} 10}{log \hspace{0.1cm} P99/P99.9} $$

Write a function, exponent.est_ratio() which takes in values for P99 and
P99.9, and returns the value of a implied by the equation above. Check
that if P99=1e6 and P99.9=1e7, your function returns an a of 2.

```{r}

exponent.est_ratio <- function(p99, p99.9){
  a <- 1 - (log(10)/log(p99/p99.9))
  return(a)
}

exponent.est_ratio(p99=1e6, p99.9=1e7)
```

# Part II

4.  Estimate a for each year in the data set, using your
    exponent.est_ratio() function. If the function was written properly,
    you should not need to use a loop. Plot your estimate of a over
    time. Do the results look reasonable? (Remember that smaller
    exponents mean more income inequality.)

    ```{r}
    result <- exponent.est_ratio(new.data$P99, new.data$P99.9)
    plot(result)

    plot(new.data$Year, result)
    ```

5.  Suppose, we consider the following.

$$ \bigg( \frac{P99.5}{P99.9} \bigg)^{-a+1} = b $$

Write a function which takes P99.5, P99.9 and a, and calculates the
left-hand side of that equation. Plot the values for each year, using
the data and your estimates of the exponent. Add a horizontal line with
vertical coordinate.

```{r}

calc_b <- function(p99.5, p99.9, a){
  b <- (p99.5/p99.9)^(-a+1)
  return(b)
}

result2 <- calc_b(new.data$P99.5, new.data$P99.9, exponent.est_ratio(new.data$P99.5,new.data$P99.9))


plot(new.data$Year, result2)
abline(h=10, ylab="", xlab="", col="red")
```

6.  Repeat the previous step $(P99/P99.5)$ with this formula. How would
    you describe this fit compared to the previous ones?

(Note: the provided formula is not the best way to estimate a, but it is
one of the simplest.)

```{r}
result3 <- calc_b(new.data$P99, new.data$P99.5, exponent.est_ratio(new.data$P99,new.data$P99.5))


plot(new.data$Year, result2)
abline(h=10, ylab="", xlab="", col="red")
```
