---
title: "Lab 2"
output: word_document
---

```{r setup, include=FALSE}
# knitr::opts_knit$set(root.dir = "<your data directory here>")
```

# Instructions

In this set of exercises, you will be devising a new grading scheme for the students in the `2009Grade.csv` data. Begin by setting your working directory in the code chunk above and loading the data into a variable called `df`.

```{r}
df <- read.csv("/Users/jasperstone/Desktop/Joseph/2023 Spring/STAT404/2009Grade.csv")
```

# Problem 1

You want to backtrack and see what grading scheme was used to initially calculate the student's grade as represented in the `Score` column of the data set. This can be accomplished using linear algebra.

1.  Select the first four rows of the `data.frame`, and store them in a variable called `first.four`.

*Hint: Recall that the first* $n$ rows of a `data.frame` can be selected using:

`df[1:n, ]`

```{r}
class(df)
first.four <- df[1:4,]
class(first.four)
first.four
```

2.  From `first.four`, put the `Mid1`, `Mid2`, `Final`, and `HW` columns into a matrix `A`. Then, put the `Score` column into a matrix called `b`.

*Hint 1: Recall that if `df` is a `data.frame` with columns named `x`, `y`, and `z`, these columns can be selected using:*

`df[ , c("x", "y", "z")]`

*Hint 2: Use the `as.matrix` function*

```{r}

A <- as.matrix(first.four[, c("Mid1","Mid2","Final","HW")])
b <- as.matrix(first.four$Score)
dim(A)
dim(b)
```

3.  This is now a system of linear equations $A\vec x = \vec b$, where $\vec x$ are the weightings. Solve for the weightings.

*Hint: Use `solve`*

```{r}
solve(A,b)
```

# Problem 2

Now that you have recovered the initial weightings, you would like to add a curve to the student's final exam scores. You will do this by:

-   Dropping the top 5 scores
-   Dropping the bottom 5 scores
-   Calculating the mean of the remaining students
-   Taking the difference between this mean and 85 to be the curve
-   Adding this curve to all of the students' final exam scores

1.  Save the final exam scores into a vector called `old.scores`. Use the `sort` function to drop the top and bottom five scores from this vector.

```{r}
old.scores<-df$Final
print(old.scores)
old.scores <- df[,"Final"]
old.scores
```

```{r}

temp <- sample(c(1:10), 10, replace=TRUE)


temp <- sort(temp)

temp <- temp[-c(1:3)]
temp <- sort(temp, decreasing = TRUE)[-c(1:3)]
class(temp)

# print(sort(df$Final))
# class(df$Final)
# old.scores <- (sort(df$Final)[-c(1,5)])
# print(".     ")
# #old.scores
# old.scores <- sort(old.scores, decreasing = TRUE)[-c(1,5)]
# print("")
# #old.scores

old.scores <- df[,"Final"]
is.data.frame(old.scores)
test <- df$Final
is.data.frame(test)

print("orig")
print(sort(df$Final))
old.scores <- (sort(df$Final))[-c(1:5)] # smallest 5 remove 
print("1st")
old.scores
old.scores <- sort(old.scores, decreasing = TRUE)[-c(1:5)] # largest 5 remove
print("2nd")
old.scores
```

2.  Calculate the curve by taking the mean of `old.scores` and subtracting it from 85. Add this curve to everyone's final exam score, and save it to a new column in the `df` called `new.final`.

*Hint 1: To add a constant value to a vector `x`, use:* `x + 5`

*Hint 2: To add a column to a `data.frame`, use:* `df$<new column name> = <vector of new values>`

```{r}
85 - mean(old.scores)
df$new.final <-  df$Final+ (85 - mean(old.scores))
df$new.final
```

# Problem 3

You will now replace the student's lowest midterm grade with their **old** final exam grade if the old final exam grade is higher.

1.  Use `cbind` to put the `Mid1` and `Mid2` columns into a matrix.

```{r}
mid1.mid2 <- cbind(df$Mid1, df$Mid2)
is.matrix(mid1.mid2)
```

2.  Find the minimum of each row and save it to a vector. Add this vector to `df` in a new column called `mid.min`.

*Hint: Use the `apply` function and `min`*

```{r}

df$mid.min <- apply(mid1.mid2, 1, min)
df$mid.min
```

3.  Find the maximum of each row and save it to a vector. Add this vector to `df` in a new column called `mid.max`.

```{r}
df$mid.max <- apply(mid1.mid2, 1, max)
df$mid.max
```

4.  Create a boolean vector that is `TRUE` for students whose lowest midterm is less than their final exam score, and is `FALSE` otherwise.

```{r}
final.better <- df$mid.min < df$Final
final.better
```

5.  How many students have a higher score on their final than on their lowest midterm?

*Hint: Try running the following line of code:*

```{r}
sum(c(TRUE, FALSE, TRUE))
```

```{r}
sum(final.better)
```

6.  Return the **old** final exam scores for the students that performed better on the final than on their lowest midterm and save it in a vector called `final.replace`.

*Hint: Recall that indexing a vector with a boolean vector (vector of `TRUE/ FALSE`) will return the original vector with the entries corresponding to `FALSE` removed*

```{r}
length(final.better)
final.replace <- df$Final[final.better]
length(final.replace)
```

7.  For students that performed better on the final than on their lowest midterm, replace the value in the `mid.min` column with the corresponding value in the `final.replace` vector.

```{r}

df$mid.min[final.better] <- final.replace 
```

# Problem 4

1.  Use these modified weightings to calculate the new final exam score using matrix multiplication:

-   `mid.min`: 10%
-   `mid.max`: 30%
-   `HW`: 10%
-   `new.final`: 50%

Save the new score in a new column called `new.score`.

*Hint: Matrix `A` can be multiplied with matrix `x` using the following syntax:* `A %*% x`

```{r}

new.score = as.matrix(df[,c("mid.min", "mid.max", "HW", "new.final")]) %*% matrix(c(0.1, 0.3, 0.1, 0.5))
df$new.score <- new.score
df$new.score
```

2.  Generate a 5 number summary of the new and old scores.

```{r}
summary(df$Score)
summary(df$new.score)
```

# Problem 5

1.  Expand on the function `grader` that assigns a letter grade based on the student's new score so that it is in accordance with the following scale:

-   A: $\geq 90$
-   B: $\geq 80$
-   C: $\geq 70$
-   D: $\geq 60$
-   F: $<60$

Try it out by running the "test data" lines of code below.

```{r}
grader <- function(score){
  
  letter.grade <- "F"
  
  if(score >= 90){
    letter.grade <- "A"
  }
  else if(score >= 80){
    letter.grade <- "B"
  }
    else if (score>=70){
      letter.grade<-"C"
    }
  else if (score>=60){
    letter.grade<-"D"
  }
  return(letter.grade)
}

# test data
test.score <- seq(0, 100, 10)
test.letter <- sapply(test.score, grader)
rbind(test.score, test.letter)
```

2.  Apply this function to the `new.score` column and save the new letters in a column called `new.grade`.

*Hint: Use the `sapply` function, as demonstrated in the `test data` section above*

```{r}
df$new.grade <- sapply(df$new.score, grader)
```

3.  Create a new column called `update`. For the *j*-th student, the value in this column is: $$\text{Old: }x_j, \text{ New}: y_j$$

where $x_j, y_j$ are the student's old letter grade and new letter grade respectively.

*Hint: Use the `paste0` function*

```{r}
df$update <- paste0("old:", df$Grade, " new:", df$new.grade)
df$update
```
