---
title: 'Lab 3'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error =TRUE)
```

***What's this about***
 
Manipulating data frames; practicing iteration; practicing re-writing code; checking
  how reliable random methods are.


## R Markdown Test

0. Open a new R Markdown file; set the output to HTML mode and "Knit". This should produce a web page with the knitting procedure executing your code blocks.


Part I -- Functions
======
1. Write an if statement by taking the input of i as 6 and printing "i is greater than 5" (since it is greater than 5).

```{r}
i <- 6
func1 <- function(i){
  if(i>5){
  print("i is greater than 5")
  }
}

func1(5)
func1(6)
```

2. Write an if/else statement for the same prompt as above, but print "i <=5" for the else loop.

```{r}
i <- 6
func2 <- function(i){
  if(i>5){
  print("i is greater than 5")
  }else{
    print("i <=5")
  }
}

func2(5)
func2(6)
```

3. Use the same concept to construct a for loop to check for i in a sequence of 1 to 10, is greater or lesser than 5.

```{r}
vec1 <- seq(1,10,1)
vec1

for(i in 1:length(vec1)){
  func2(vec1[i])
}
```

4. Using positional indexes- for string 'TX','CA','NY', print the prefix "The state is ".

```{r}
states <- c('TX','CA','NY')
states

for(i in 1:length(states)){
  print(paste("The state is ", states[i]))
}

# Is this python way?
#for(i in states){
#  print(paste("The state is ", states))
#}
```

5. Take the inputs- y1=c(3,20,5,3,8);y2=log(y1); Use positional indexes to save the result of y1*y2, but don't save when it is greater than 50.
```{r}
y1=c(3,20,5,3,8)
y2=log(y1)
result.vec <- vector()


for(i in 1:length(y1)){
  y1.y2 = y1[i]*y2[i]
  if(y1.y2>50){
    next
  }
  result.vec = c(result.vec, y1.y2)
}

y1*y2
result.vec
```

6. Write a for loop to draw 10 random samples from a normal distribution with mean 6 and standard deviation 2 (save it as y). Check if y is greater than 6 or not.
```{r}
y <- rnorm(10, 6,2)
y
for(i in 1:length(y)){
  if(y[i]>6){
    print(paste("y[", i,"] is", y[i], " which is greater than 6"))
  }else{
    print(paste("y[", i,"] is", y[i], " which is not greater than 6"))
  }
}
```

7. Use the prompt from the question above, create a new vector that only saves the y values which are >6.
```{r}
y <- rnorm(10, 6,2)
y
y.vec <- vector()
for(i in 1:length(y)){
  if(y[i]>6){
    y.vec <- c(y.vec, y[i])
  }else{
    next
  }
}
y.vec
```

Part II -- Reproducibility and Functions
==========================================

Below is a piece of code that optimizes the factory's output (roughly) given the available resources, using a `repeat` loop.  It's embedded in a function to make it easier for you to run.

```{r}
factory.function <- function (cars.output, trucks.output) {
  factory <- matrix(c(40,1,60,3),nrow=2,
    dimnames=list(c("labor","steel"),c("cars","trucks")))
  available <- c(1600,70); names(available) <- rownames(factory)
  slack <- c(8,1); names(slack) <- rownames(factory)
  output <- c(cars.output, trucks.output); names(output) <- colnames(factory)

  passes <- 0 # How many times have we  been around the loop?
  repeat {
     passes <- passes + 1
     needed <- factory %*% output # What do we need for that output level?
     # If we're not using too much, and are within the slack, we're done
     if (all(needed <= available) &&
         all((available - needed) <= slack)) {
       break()
     }
     # If we're using too much of everything, cut back by 10%
     if (all(needed > available)) {
       output <- output * 0.9
       next()
     }
     # If we're using too little of everything, increase by 10%
     if (all(needed < available)) {
       output <- output * 1.1
       next()
     }
     # If we're using too much of some resources but not others, randomly
     # tweak the plan by up to 10%
      # runif == Random number, UNIFormly distributed, not "run if"
     output <- output * (1+runif(length(output),min=-0.1,max=0.1))
  }

  return(output)
}
```

5. Run the function above with the command `factory.function(1,1)` to obtain a default output value, starting from a very low initial planned output. What is the final output capacity obtained?
```{r}
factory.function(1,1)
```


6. Repeat this four more times to obtain new output values. Do these answers differ from each other? If so why? If not, why not?
```{r}
factory.function(1,1)
factory.function(1,1)
factory.function(1,1)

```

In the function we have a random uniform distribution. 


7. Right now, the number of `passes` is a value held within the function itself and not shared. Change the code so that the number of `passes` will be returned at the end of the function, as well as the final `output`. 
```{r}
factory.function2 <- function (cars.output, trucks.output) {
  factory <- matrix(c(40,1,60,3),nrow=2,
    dimnames=list(c("labor","steel"),c("cars","trucks")))
  available <- c(1600,70); names(available) <- rownames(factory)
  slack <- c(8,1); names(slack) <- rownames(factory)
  output <- c(cars.output, trucks.output); names(output) <- colnames(factory)

  passes <- 0 # How many times have we  been around the loop?
  repeat {
     passes <- passes + 1
     needed <- factory %*% output # What do we need for that output level?
     # If we're not using too much, and are within the slack, we're done
     if (all(needed <= available) &&
         all((available - needed) <= slack)) {
       break()
     }
     # If we're using too much of everything, cut back by 10%
     if (all(needed > available)) {
       output <- output * 0.9
       next()
     }
     # If we're using too little of everything, increase by 10%
     if (all(needed < available)) {
       output <- output * 1.1
       next()
     }
     # If we're using too much of some resources but not others, randomly
     # tweak the plan by up to 10%
      # runif == Random number, UNIFormly distributed, not "run if"
     output <- output * (1+runif(length(output),min=-0.1,max=0.1))
  }
  output <- c(output, passes)
  #
  names(output) <- c(colnames(factory), "passes")
  return(output)
}

factory.function2(1,1)
```


8. What is the final output plan (`output`) if we want to make 30 cars and 20 trucks?  What is the final demand for resources (`needed`)?  How many iterations did it take to converge (`passes`)? For all but `output` you will need to either print this message out deliberately, or return an object that contains all the quantities you want.
```{r}
factory.function3 <- function (cars.output, trucks.output) {
  factory <- matrix(c(40,1,60,3),nrow=2,
    dimnames=list(c("labor","steel"),c("cars","trucks")))
  available <- c(1600,70); names(available) <- rownames(factory)
  slack <- c(8,1); names(slack) <- rownames(factory)
  output <- c(cars.output, trucks.output); names(output) <- colnames(factory)

  passes <- 0 # How many times have we  been around the loop?
  repeat {
     passes <- passes + 1
     needed <- factory %*% output # What do we need for that output level?
     # If we're not using too much, and are within the slack, we're done
     if (all(needed <= available) &&
         all((available - needed) <= slack)) {
       break()
     }
     # If we're using too much of everything, cut back by 10%
     if (all(needed > available)) {
       output <- output * 0.9
       next()
     }
     # If we're using too little of everything, increase by 10%
     if (all(needed < available)) {
       output <- output * 1.1
       next()
     }
     # If we're using too much of some resources but not others, randomly
     # tweak the plan by up to 10%
      # runif == Random number, UNIFormly distributed, not "run if"
     output <- output * (1+runif(length(output),min=-0.1,max=0.1))
  }
  output <- c(output, passes, needed)
  #
  names(output) <- c(colnames(factory), "passes","labor","steel")
  return(output)
}

factory.function3(30,20)
```

