---
title: "Lab 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error =TRUE)
```

***What's this about***

Manipulating data objects; using the built-in functions,
  doing numerical calculations, and basic plots; reinforcing core probabilistic
  ideas.


## R Markdown Test

0. Open a new R Markdown file; set the output to HTML mode and "Knit". This should produce a web page with the knitting procedure executing your code blocks.

```{r}
## For reproducibility --- don't change this!
set.seed(01202023)
```

# The Exponential Distribution

The exponential distribution, $\mathrm{Exp}(\lambda)$ is defined by its cumulative distribution function

$$ F(x) = 1-e^{-\lambda x} $$

The R function `rexp` generates random variables with an exponential
distribution.

```
rexp(n=10, rate=5)
```

produces 10 exponentially-distributed numbers with rate ($\lambda$) of 5.  If
the second argument is omitted, the default rate is 1; this is the ``standard
exponential distribution''.

# Some simple manipulations and simple plots

1. Generate 200 random values from the standard exponential distribution and store them in a vector `exp.draws.1`.  Find the mean and standard deviation of `exp.draws.1`.

```{r}
exp.draws.1 <- rexp(n=200)
mean(exp.draws.1)
sd(exp.draws.1)
```


2. Repeat, but change the rate to 0.1, 0.5, 5 and 10, storing the results in vectors called  `exp.draws.0.1`,  `exp.draws.0.5`,  `exp.draws.5` and  `exp.draws.10`. 

```{r}
exp.draws.0.1 = rexp(n=200, rate=0.1)
exp.draws.0.5 = rexp(n=200, rate=0.5)
exp.draws.5 = rexp(n=200, rate=5)
exp.draws.10 = rexp(n=200, rate=10)
```


3. The function `plot()` is the generic function in R for the visual display of data. `hist()` is a function that takes in and bins data as a side effect. To use this function, we must first specify what we'd like to plot.
    a. Use the `hist()` function to produce a histogram of your standard exponential distribution. 
    b. Use `plot()` with this vector to display the random values from your standard distribution in order.
    c. Now, use `plot()` with two arguments -- any two of your other stored random value vectors -- to create a scatterplot of the two vectors against each other.

```{r}
hist(exp.draws.1)
```
```{r}
plot(exp.draws.1)
```

```{r}
plot(x=exp.draws.1, y=exp.draws.5)
```

4. We'd now like to compare the properties of each of our vectors. Begin by creating a vector of the means of each of our five distributions in the order we created them and saving this to a variable name of your choice. Using this and other similar vectors, create the following scatterplots:
    a. The five means versus the five rates used to generate the distribution.
    b. The standard deviations versus the rates.
    c. The means versus the standard deviations.

For each plot, explain in words what's going on.

```{r}
exp.draws.1.mean = mean(exp.draws.1)
exp.draws.0.1.mean =  mean(exp.draws.0.1)
exp.draws.0.5.mean = mean(exp.draws.0.5)
exp.draws.5.mean = mean(exp.draws.5)
exp.draws.10.mean = mean(exp.draws.10)

```

```{r}
exp.draws.means = c(exp.draws.0.1.mean,exp.draws.0.5.mean,exp.draws.1.mean,exp.draws.5.mean,exp.draws.10.mean)
rates = c(0.1, 0.5, 1, 5, 10)
plot(x=rates, y=exp.draws.means)


```
```{r}
exp.draws.1.sd = sd(exp.draws.1)
exp.draws.0.1.sd =  sd(exp.draws.0.1)
exp.draws.0.5.sd = sd(exp.draws.0.5)
exp.draws.5.sd = sd(exp.draws.5)
exp.draws.10.sd = sd(exp.draws.10)
exp.draws.sds = c(exp.draws.0.1.sd, exp.draws.0.5.sd, exp.draws.1.sd, exp.draws.5.sd, exp.draws.10.sd)
plot(x=rates, y=exp.draws.sds)


```
```{r}
plot(x=exp.draws.means, y=exp.draws.sds)
```
**Challenge**: for each plot, add a curve that corresponds to the relationships you'd expect to see in the theoretical population (i.e., with an infinite amount of draws, rather than just 200 draws).

```{r}
plot(x=rates, y=exp.draws.means)
plot(x=rates, y=exp.draws.sds)
plot(x=exp.draws.means, y=exp.draws.sds)



```

# Let's go big

5. R's capacity for data and computation is large to what was available many years ago. 
    a. To show this, generate 1.1 million numbers from the standard exponential distribution and store them in a vector called `big.exp.draws.1`. Calculate the mean and standard deviation.
    b. Plot a histogram of `big.exp.draws.1`.  Does it match the function $1-e^{-x}$?  Should it? 
    c. Find the mean of all of the entries in `big.exp.draws.1` which are strictly greater than 1. You may need to first create a new vector to identify which elements satisfy this.
    d. Create a matrix, `big.exp.draws.1.mat`, containing the the values in 
`big.exp.draws.1`, with 1100 rows and 1000 columns. Use this matrix as the input to the `hist()` function. Compare the result with the histogram in part b.
    e. Calculate the mean of the 371st column of `big.exp.draws.1.mat`.
    f. Now, find the means of all 1000 columns of `big.exp.draws.1.mat` simultaneously. Plot the histogram of column means.  Explain why its shape does not match the histogram in problem 5b).
    g. Take the square of each number in `big.exp.draws.1`, and find the mean of this new vector.  Explain this in terms of the mean and standard deviation of `big.exp.draws.1`. ***Hint:*** think carefully about the formula R uses to calculate the standard deviation.

```{r}
big.exp.draws.1 <- rexp(n=1100000)
mean(big.exp.draws.1)
sd(big.exp.draws.1)
```


```{r}
hist(big.exp.draws.1)
```

```{r}

hist(big.exp.draws.1)
xgrid <- seq(0,15, len=10000)
yval <- exp(-xgrid) # density.  # exp is not exp dist. here if we want to get the pdf for exp dist with lambda 1.
hist(big.exp.draws.1, freq=FALSE)
lines(xgrid, yval)

```

```{r}
big.exp.draws.1.subset <- big.exp.draws.1[big.exp.draws.1 > 1]
mean(big.exp.draws.1.subset)
```

d.  Create a matrix, `big.exp.draws.1.mat`, containing the the values in `big.exp.draws.1`, with 1100 rows and 1000 columns. Use this matrix as the input to the `hist()` function. Compare the result with the histogram in part b.

```{r}
big.exp.draws.1.mat <- matrix(data = big.exp.draws.1, nrow=1100, ncol=1000)
hist(big.exp.draws.1.mat)
```

e.  Calculate the mean of the 371st column of `big.exp.draws.1.mat`.
```{r}
mean(big.exp.draws.1.mat[,371])
```

f.  Now, find the means of all 1000 columns of `big.exp.draws.1.mat` simultaneously. Plot the histogram of column means. Explain why its shape does not match the histogram in problem 5b).

```{r}
col.mean.vec <- colMeans(big.exp.draws.1.mat)
hist(col.mean.vec)
```


g.  Take the square of each number in `big.exp.draws.1`, and find the mean of this new vector. Explain this in terms of the mean and standard deviation of `big.exp.draws.1`. ***Hint:*** think carefully about the formula R uses to calculate the standard deviation.

```{r}
mean((big.exp.draws.1)^2)

# var(x) = E(x^2) - [E(x)]^2
sd(big.exp.draws.1)^2 + mean(big.exp.draws.1)^2

```


# Now let's go really big

6. Let’s push R’s computational engine a little harder. Design an expression to generate 100 million numbers from $\mathrm{Exp}(5)$ to be saved in a vector called `huge.exp.draws`, but do not evaluate this command yet. 
    a. Then ask the person next to you to name three of Justin Bieber’s songs and simultaneously evaluate your R command that defines `huge.exp.draws`. Which finished first, R or them? 

```{r}
huge.exp.draws = rexp(n=100000000, rate = 5)
```

    b. Calculate the mean and standard deviation of `huge.exp.draws`. Are they close to what you'd expect? Did R take longer to compute these, or to generate `huge.exp.draws` in the first place?
```{r}
mean(huge.exp.draws)
sd(huge.exp.draws)

system.time(rexp(n = 1e8, rate = 5))
system.time(mean(huge.exp.draws))
system.time(sd(huge.exp.draws))
```

    c. Calculate the median of `huge.exp.draws`. Did this median calculation take longer than calculating the mean? Is this surprising?
```{r}
median(huge.exp.draws)

system.time(median(huge.exp.draws))
```

    d. Calculate the exponential of the median of the logs of `huge.exp.draws`, in one line of code. Did this take longer than the median calculation applied to `huge.exp.draws` directly? Is this surprising?
```{r}
exp(median(log(huge.exp.draws)))

system.time(exp(median(log(huge.exp.draws))))
```

    e. Plot a histogram of of `huge.exp.draws`, again with a large setting of the `breaks` argument (e.g., `breaks=100`). Describe what you see; is this different from before, when we had 3 million draws? **Challenge**: Is this surprising? What distribution is this?
```{r}
hist(huge.exp.draws, breaks = 100)

```

