---
title: 'Notes 3: Data Frames and Control'
author: "STAT 404: Statistical Computing"

# output:
#  revealjs::revealjs_presentation:
#    css: slides_styles.css
#    center: false
#    transition: slide
#    #self_contained: false
#    #reveal_plugins: ["chalkboard"]
#    reveal_options:
#      slideNumber: true
#    #  navigationMode: 'linear'

# for course output; files in folder output
output:
  html_document: default
  pdf_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_dir = "output", output_format = "all") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error =TRUE)
```

## Recap 

1. What attribute defines an S3 object on a base type? What are some S3 objects we have discussed?

  Ans: If it has a 'class' attribute, it's S3. Data frame has class 'data.frame'

2. .Rmd demo: Open a new .Rmd file and practice the following three ways to insert a new code chunk

- The keyboard shortcut Cmd + Option + I / Ctrl + Alt + I.

```{r}

```

- The “Insert” button icon in the editor toolbar.
```{r}

```

- By manually typing the chunk delimiters ```` ```{r} ```` and ```` ``` ````.

```{r}
```
3. What do the chunk options `eval` and `echo` control?

```{r eval=FALSE}

```


  Ans: `eval` : run code in chunk (True - evaluate, False - not evaluate)
        `echo` : display mode (True - show, False - not show)
---

## Outline

- Making and working with data frames
    - Subsetting
    - Adding new variables (columns)
    - Removing variables (columns)

---

## In our last thrilling episode
- Atomic vectors: series of values all of the same type  
  e.g., `v[5]`, `v["name"]`
- Arrays: multi-dimensional generalization of atomic vectors
  e.g., `a[5,6,2]`, `a[,6,]`, `a["rowname", "colname", "layername"]`
- Matrices: special 2D arrays with matrix math  
  e.g., `m[5,6]`, `m[,6]`, `m[,"colname"]`
- Lists: vector of values of mixed types  
  e.g., `l[[3]]`, `l$name`
- Data frames: list with **data.frame** class attribute; matrix and list indexing work

---

## Data frames, encore
- 2D tables of data
- Each case/observation is a row
- Each variable/feature is a column
- Variables can be of any type (numbers, text, Booleans, ...)
- Both rows and columns can get names

---

## Creating an example data frame

Use `data.frame()`, similar to how we create lists with `list()`

```{r}
my.df = data.frame(nums=seq(0.1,0.6,by=0.1), chars=letters[1:6], 
                   bools=sample(c(TRUE,FALSE), 6, replace=TRUE))
my.df

# Note, a list can have different lengths for different elements!
my.list = list(nums=seq(0.1,0.6,by=0.1), chars=letters[1:12], 
               bools=sample(c(TRUE,FALSE), 6, replace=TRUE))
my.list
```

---

## Indexing a data frame

- By rows/columns: similar to how we index matrices
- By columns only: similar to how we index lists

```{r}
my.df[,1] # Also works for a matrix 
my.df[,"nums"] # Also works for a matrix
my.df$nums # Doesn't work for a matrix, but works for a list
my.df$chars # Note: this one has been converted into a factor data type

is.factor(my.df$chars)

as.character(my.df$chars) # Converting it back to a character data type
is.factor(as.character(my.df$chars))


```

---

## Creating a data frame from a matrix

Often times it's helpful to start with a matrix, and add columns (of different data types) to make it a data frame

```{r}
class(state.x77) # Built-in matrix of states data, 50 states x 8 variables
head(state.x77) 
class(state.region) # Factor of regions for the 50 states
head(state.region)
class(state.division) # Factor of divisions for the 50 states
head(state.division) 
```

---

```{r}
# Combine these into a data frame with 50 rows and 10 columns
state.df = data.frame(state.x77, Region=state.region, Division=state.division)
class(state.df)
head(state.df) # Note that the first 8 columns name carried over from state.x77
```
`data.frame()` is combining a pre-existing matrix (`state.x77`) and two vectors of qualitative categorical variables (called **factors**; `state.region`, `state.division`)

Column names are preserved or guessed if not explicitly set

---

```{r}
colnames(state.df)
state.df[1,]
```

---

## Data frame access
By row and column index
```{r}
state.df[49,3]
```

By row and column names
```{r}
state.df["Wisconsin","Illiteracy"]
```

---

## Data frame access (cont'd)
All of a row:
```{r}
state.df["Wisconsin",]
```
<small>Exercise: what class is `state.df["Wisconsin",]`?</small>

```{r}
is.list(state.df["Wisconsin",])
is.data.frame(state.df["Wisconsin",])
class(state.df["Wisconsin",])
```

---

## Data frame access (cont'd.)
All of a column:
```{r}
head(state.df[,3])
head(state.df[,"Illiteracy"])
head(state.df$Illiteracy)
```

---

## Data frame access (cont'd.)
Rows matching a condition:
```{r}
state.df[state.df$Division=="New England", "Illiteracy"]
state.df[state.df$Region=="South", "Illiteracy"]
```

---

## Adding columns to a data frame

To add columns: we can either use `data.frame()`, or directly define a new named column

```{r}
# First way: use data.frame() to concatenate on a new column
state.df = data.frame(state.df, Cool=sample(c(T,F), nrow(state.df), rep=TRUE))
head(state.df, 4)

# Second way: just directly define a new named column
state.df$Score = sample(1:100, nrow(state.df), replace=TRUE)
head(state.df, 4)
```

---

## Deleting columns from a data frame

To delete columns: we can either use negative integer indexing, or set a column to `NULL`

```{r}
# First way: use negative integer indexing
ncol(state.df) ## this is 12, therefore remove the 12th col
state.df = state.df[,-ncol(state.df)]
head(state.df, 4)

# Second way: just directly set a column to NULL
state.df$Cool = NULL
head(state.df, 4)
```

---

## Reminder: Boolean indexing

With matrices or data frames, we'll often want to access a subset of the rows corresponding to some condition. You already know how to do this, with Boolean indexing

```{r}
# Compare the averages of the Frost column between states in New England and
# Pacific divisions
t <- state.df$Division == "New England"
t
mean(state.df[(state.df$Division == "New England"), "Frost"]) 
mean(state.df[(state.df$Division == "Pacific"), "Frost"]) 
```
> What is the average of Frost for the division that contains Texas?

```{r}

texas.division <- state.df["Texas","Division"]
texas.division
division <- state.df$Division==texas.division
state.df$Frost
mean(state.df[division,"Frost"])
```

---

## `subset()`

The `subset()` function provides a convenient alternative way of accessing rows for data frames

```{r}
# Using subset(), we can just use the column names directly (i.e., no need for
# using $)
state.df.ne.1 = subset(state.df, Division == "New England")
state.df.ne.1

state.df$Division == "New England"
t <- state.df[state.df$Division == "New England",]
t

# Get same thing by extracting the appropriate rows manually
state.df.ne.2 = state.df[state.df$Division == "New England", ]
state.df.ne.2
all(state.df.ne.1 == state.df.ne.2)

# Same calculation as in the last slide, using subset()
mean(subset(state.df, Division == "New England")$Frost)
mean(subset(state.df, Division == "Pacific")$Frost) # Wimps
```

---

## Replacing values
Parts or all of the data frame can be assigned to:
```{r}
summary(state.df$HS.Grad)
state.df$HS.Grad <- state.df$HS.Grad/100
summary(state.df$HS.Grad)
state.df$HS.Grad <- 100*state.df$HS.Grad
```

---

## `with()`

The `with()` function provides a way of expressing operations by column names only.

What percentage of literate adults graduated high school?
```{r}
head(100*(state.df$HS.Grad/(100-state.df$Illiteracy)))
# highschool grad / literacy
```

`with()` takes a data frame and evaluates an expression "inside" it:
```{r}
with(state.df, head(100*(HS.Grad/(100-Illiteracy))))
```
(so you don't have to type `state.df$xyz`)

---

## Data arguments
Lots of functions take `data` arguments, and look variables up in
that data frame:
```{r fig.height=5, fig.width=5}
plot(Illiteracy~Frost, data=state.df)
```

---

## Summary

- Data frames are a representation of the "classic" data table in R: rows are observations/cases, columns are variables/features
- Each column can be a different data type (but must be the same length)
- `subset()`: function for extracting rows of a data frame meeting a condition 
- `with()`: function for operating on data frame columns without indexing
