---
title: 'Notes 2: More Data Structures'
author: "STAT 404: Statistical Computing"

# output:
#   revealjs::revealjs_presentation:
#     css: slides_styles.css
#     center: false
#     transition: slide
#     #self_contained: false
#     #reveal_plugins: ["chalkboard"]
#     reveal_options:
#       slideNumber: true
#     #  navigationMode: 'linear'

## for course output; files in folder output
#output:
#  html_document: default
#  pdf_document: default
knit: (function(inputFile, encoding) {
 rmarkdown::render(inputFile, encoding = encoding,
 output_dir = "output", output_format = "all") })
---

## Outline

- Arrays
- Matrices
- Lists
- Data frames
- Structures of structures

---

## Five basic data structures

R has five most oftenly used data structures:

* Atomic vector: 1d, homogeneous
* Matrices: 2d, homogeneous
* Arrays: $n$-d, homogeneous
* Lists: 1d, heterogeneous
* Data frames: 2d, heterogeneous

---

## Vector structures, starting with arrays

Many data structures in R are made by adding bells and whistles to vectors, i.e., they are "vector structures".

An **array** is an atomic vector with a **dimension** attribute.
```{r}
x <- c(7, 8, 10, 45)
dim(x) <- c(2,2)
x
x.arr <- array(x,dim=c(2,2))
x.arr
```

- `dim` says how many rows and columns; filled by columns

- Can have 3, 4, ... dimensional arrays; `dim` is a vector of arbitrary length

---

Some properties of the array:
```{r}
dim(x.arr)
is.vector(x.arr)
is.array(x.arr)
typeof(x.arr)
```

---

```{r}
typeof(x.arr)
str(x.arr)
attributes(x.arr)
```

`typeof()` returns the type of the _elements_

`str()` gives the structure: here, a numeric array, with two dimensions, both indexed 1--2, and then the actual numbers

Exercise: try all these with `x`

---

## Accessing and operating on arrays

Can access a 2-D array either by pairs of indices or by the underlying vector:

```{r}
x.arr[1,2]
x.arr[3]
```

---

```{r}
x.arr[,2] # Omitting an index means "all of it"
x.arr[c(1:2),2]
x.arr[,2,drop=FALSE]
```

> Recall that attributes may not persist. The optional third argument `drop=FALSE` ensures that the dimension attribute persists, i.e., the results are still an array, not a vector.

---

## Functions on arrays

Many functions applied to an array will just boil things down to the underlying vector, *unless* the function is set up to handle arrays specially:
```{r}
which(x.arr > 9) # recall that x.arr is a 2-D array
```

---

And there are several functions/operators that _do_ preserve array structure:
```{r}
y <- -x
y.arr <- array(y,dim=c(2,2))
y.arr + x.arr
```

Others specifically act on each row or column of the array separately:
```{r}
rowSums(x.arr)
```

<small>(We will see a lot more of this idea)</small>

---

## Example: Price of houses in PA

Census data for California and Pennsylvania on housing prices,
by Census "tract"
```{r}
#calif_penn <- read.csv("data/calif_penn_2011.csv") # set the right relative path!
calif_penn <- read.csv("~/Desktop/Joseph/2023 Spring/STAT404/calif_penn_2011.csv") # set the right relative path!

penn <- calif_penn[calif_penn[,"STATEFP"]==42,]
coefficients(lm(Median_house_value ~ Median_household_income, data=penn))
```
Fit a simple linear model, predicting median house price from median household income

---

Rows 24--425 of `penn` are tracts for Allegheny county

Row 24 has a median income of \$14,719; Is the regression line higher or lower than the actual median house value \$34,100?
```{r}
34100 < -26206.564 + 3.651*14719
```

Row 25 has income \$48,102 and house price \$155,900
```{r}
155900 < -26206.564 + 3.651*48102
```

What about Row 26?

---

We _could_ just keep plugging in numbers like this, but that's

- boring and repetitive
- error-prone (what if I forget to change the median income, or drop a minus sign from the intercept?)
- obscure if we come back to our work later (what _are_ these numbers?)

---

## Use variables and names

```{r}
penn.coefs <- coefficients(lm(Median_house_value ~ Median_household_income,
                              data=penn))
penn.coefs
```

```{r}
allegheny.rows <- 24:425
allegheny.medinc <- penn[allegheny.rows,"Median_household_income"]
allegheny.values <- penn[allegheny.rows,"Median_house_value"]
allegheny.fitted <- penn.coefs["(Intercept)"] +
  penn.coefs["Median_household_income"]*allegheny.medinc
```

---

```{r}
plot(x=allegheny.fitted, y=allegheny.values,
     xlab="Model-predicted median house values",
     ylab="Actual median house values",
     xlim=c(0,5e5),ylim=c(0,5e5))
abline(a=0,b=1,col="grey")
```

---

## Running example: resource allocation ("mathematical programming")

Factory makes cars and trucks, using labor and steel

- a car takes 40 hours of labor and 1 ton of steel
- a truck takes 60 hours and 3 tons of steel
- resources: 1600 hours of labor and 70 tons of steel each week

Let $x$ and $y$ be the number of cars and trucks respectively.
The above constraint can be represented by 
$$
\begin{bmatrix}
40 & 60\\
1 & 3
\end{bmatrix}
\begin{bmatrix}
x \\ y
\end{bmatrix}
\le
\begin{bmatrix}
1600 \\ 70
\end{bmatrix}
$$
($\le$ is interpreted in an element-wise sense)

To use all the resources, we may make it "equality".

---

## Matrices

In R, a matrix is a specialization of a 2D array

```{r}
factory <- matrix(c(40,1,60,3),nrow=2)
is.array(factory)
is.matrix(factory)
```

- Could also specify `ncol`, and/or `byrow=TRUE` to fill by rows
- Element-wise operations with the usual arithmetic and comparison operators (e.g., `factory/3`)

---

## Matrix multiplication

Has its own special operator `%*%`

```{r}
six.sevens <- matrix(rep(7,6),ncol=3)
six.sevens
factory %*% six.sevens # [2x2] * [2x3]
#six.sevens %*% factory
```

<small>(What happens if you try `six.sevens %*% factory`?)</small>

---

## Multiplying matrices and vectors

Numeric vectors can act like proper vectors:
```{r}
output <- c(10,20)
factory %*% output
output %*% factory
```
<small>(R silently casts the vector as either a 1-row or a 1-column matrix)</small>

---

## Matrix operators

Transpose:
```{r}
t(factory)
```

Determinant:
```{r}
det(factory)
```

---

## The diagonal

The `diag()` function can extract the diagonal entries of a matrix:
```{r}
diag(factory)
```
It can also _change_ the diagonal:
```{r}
diag(factory) <- c(35,4)
factory
```
Re-set it for later:
```{r}
diag(factory) <- c(40,3)
```

---

## Creating a diagonal or identity matrix

```{r}
diag(c(3,4))
diag(2)
```
Note: the `diag(2)` is not a 1-by-1 diagonal matrix with a diagonal element 2)

---

## Inverting a matrix

```{r}
solve(factory)
factory %*% solve(factory)
```

---

## Why's it called "solve"" anyway?

Solving the linear system $\mathbf{A} \mathbf{x} = \mathbf{b}$ for $\mathbf{x}$:

```{r}
available <- c(1600,70)
solve(factory,available)
factory %*% solve(factory,available)
```

---

## Names in matrices

- We can name either rows or columns or both, with `rownames()` and `colnames()`
- These are just character vectors, and we use them just like we do `names()` for vectors
- Names help us understand what we're working with
- Names can be used to coordinate different objects

---

```{r}
rownames(factory) <- c("labor","steel")
colnames(factory) <- c("cars","trucks")
factory
available <- c(1600,70)
names(available) <- rownames(factory)
```

---

```{r}
output <- c(20,10)
names(output) <- c("trucks","cars")
factory %*% output # But we've got cars and trucks mixed up! (순서가 바뀜)
factory %*% output[colnames(factory)]
all(factory %*% output[colnames(factory)] <= available[rownames(factory)])
###
output
colnames(factory)
output[colnames(factory)]
```
Note: Last lines don't have to change if we add motorcycles as output or rubber and glass as inputs (abstraction again)

---

## Doing the same thing to each row or column

Take the mean: `rowMeans()`, `colMeans()`: input is matrix,
output is vector.  Also `rowSums()`, etc.

`summary()`: vector-style summary of column

```{r}
factory
colMeans(factory)
summary(factory)
```

---

`apply()`, takes 3 arguments:

- the array or matrix
- then 1 for rows and 2 for columns, and so on.
- then a name of the function to apply to each

```{r}
rowMeans(factory)
apply(factory,1,mean)

##
apply(factory,1,sd)
```

<small>(What would `apply(factory,1,sd)` do?)</small>

---

## Lists

Sequence of values, _not_ necessarily all of the same type

```{r}
my.distribution <- list("exponential", 7, FALSE)
my.distribution
```

Most of what you can do with vectors you can also do with lists

---

## Accessing pieces of lists

* Can use `[ ]` as with vectors 
* or use `[[ ]]`, but only with a single index  
* `[[ ]]` drops names and structures, `[ ]` does not
```{r}
my.distribution[2]
str(my.distribution[2])
```

---

```{r}
#
my.distribution

my.distribution[c(1,2)]
my.distribution[[2]]
my.distribution[[2]]^2
```
<small>(What happens if we try `my.distribution[2]^2`?)</small>
<small>(What happens if we try `[[ ]]` on a vector? Try on a named vector)</small>

---

## Expanding and contracting lists

Add to lists with `c()` (also works with vectors):
```{r}
my.distribution <- c(my.distribution,7)
my.distribution
```

---

Chop off the end of a list by setting the length to something smaller (also works with vectors):
```{r}
length(my.distribution)
length(my.distribution) <- 3
my.distribution
```

---

## Naming list elements

We can name the elements of a list
```{r}
names(my.distribution) <- c("family","mean","is.symmetric")
my.distribution
my.distribution[["family"]]
my.distribution["family"]
```

---

Lists have a special short-cut way of using names, `$`
```{r}
my.distribution[["family"]]
my.distribution$family
```

---

## Names in lists (cont'd.)

Creating a list with names:
```{r}
another.distribution <- list(family="gaussian",mean=7,sd=1,is.symmetric=TRUE)
another.distribution
```

Adding named elements:
```{r}
my.distribution
my.distribution$was.estimated <- FALSE
my.distribution[["last.updated"]] <- "2011-08-30"
my.distribution
```

Removing a named list element, by assigning it the value `NULL`:
```{r}
my.distribution$was.estimated <- NULL
my.distribution
```

---

## Key-value pairs

Lists give us a way to store and look up data by _name_, rather than by _position_

A really useful programming concept with many names: **key-value pairs**, **dictionaries**, **associative arrays**, **hashes**

If all our distributions have components named `family`, we can look that up by name, without caring where it is in the list

---

## Data frames

Data frame = the classic data table, $n$ rows for cases, $p$ columns for variables

Lots of the really-statistical parts of R presume data frames

Not just a matrix because *columns can have different types*

Many matrix functions also work for data frames (`rowSums()`, `summary()`, `apply()`)

<small>(But no matrix multiplying data frames, even if all columns are numeric)</small>


---

```{r}
a.matrix <- matrix(c(35,8,10,4),nrow=2)
colnames(a.matrix) <- c("v1","v2")
a.matrix
a.matrix[,"v1"]  # Try a.matrix$v1 and see what happens

##
a.matrix$v1 #and see what happens
###
typeof(a.matrix)
a.matrix[colnames("v1")] 
a.matrix["v1"] 
```

---

```{r}
a.data.frame <- data.frame(a.matrix,logicals=c(TRUE,FALSE))
a.data.frame
a.data.frame$v1
a.data.frame[,"v1"]
a.data.frame[1,]
```


---

```{r}
colMeans(a.data.frame)
```


What happens if we attempt to mix different types in a *matrix* by `matrix(c(3, 8, 10, 4, TRUE, FALSE), nrow=2)`?

How about `matrix(c(3, 8, 10, 4, "yes", "no"), nrow=2)`?
```{r}
matrix(c(3, 8, 10, 4, TRUE, FALSE), nrow=2)
matrix(c(3, 8, 10, 4, "yes", "no"), nrow=2)
```

---

## Adding rows and columns

We can add rows or columns to an array or data-frame with `rbind()` and `cbind()`, but be careful about forced type conversions
```{r}
rbind(a.data.frame,list(v1=-3,v2=-5,logicals=TRUE))
rbind(a.data.frame,c(3,4,6))
```

---

## Structures of Structures

So far, every list element has been a single data value

List elements can be other data structures, e.g., vectors and matrices:
```{r}
plan <- list(factory=factory, available=available, output=output)
plan$output

#
plan
```

Internally, a data frame is basically a list of vectors (all of the same length)

---

## Structures of Structures (cont'd.)

List elements can even be other lists  
which may contain other data structures  
including other lists  
which may contain other data structures...

This **nesting** lets us build arbitrarily complicated data structures from the basic ones

Most complicated objects are (usually) lists of data structures

---

## Example: Eigenstuff

`eigen()` finds eigenvalues and eigenvectors of a matrix  
Returns a list of a vector (the eigenvalues) and a matrix (the eigenvectors)

```{r}
eigen(factory)
class(eigen(factory))
```
---

With complicated objects, you can access parts of parts (of parts...)
```{r}

#
factory

factory %*% eigen(factory)$vectors[,2] # matrix * eigen vector
eigen(factory)$values[2] * eigen(factory)$vectors[,2] # eigen value * eigen vector
eigen(factory)$values[2]

#
eigen(factory)
eigen(factory)[1]
eigen(factory)[2]
(eigen(factory)[1])[2]
eigen(factory)[[1]]
typeof(eigen(factory)[[1]])
eigen(factory)$values

eigen(factory)[[1]][[2]] # NOT [[1,2]]

```

---

## Summary

- Arrays add multi-dimensional structure to vectors
- Matrices act like you'd hope they would
- Lists let us combine different types of data
- Data frames are hybrids of matrices and lists, for classic tabular data
- Recursion lets us build complicated data structures out of the simpler ones
