---
title: "STAT404_review"
output: html_document
date: "2023-01-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1/20

# Type coercion

-   R will **coerce** values to be the same type
-   R follows a fixed coercion order until all values are the same:

character $\leftarrow$ double $\leftarrow$ integer $\leftarrow$ logical

# Testing types

-   Inf is infinity.
-   NaN means 'Not a Number'

cf  - a non-zero number divided by zero creates infinity\
- NA: 'Not Available' which is not a number as well

```{r}
pi / 0 ## = Inf a non-zero number divided by zero creates infinity
0 / 0  ## =  NaN

1/0 + 1/0 # Inf
1/0 - 1/0 # NaN

is.numeric(7/0)
is.na(7/0)
typeof(7/0)

is.numeric(0/0)
is.na(0/0)
typeof(0/0)
```

## 1/23

# Peculiarites
Usually better to use all.equal() than exact comparison
```{r}
(0.5 - 0.3) == (0.3 - 0.1)
all.equal(0.5-0.3, 0.3-0.1)

is.integer(7)
is.integer(as.integer(7))
typeof(2)
typeof(2L)
```

# Matrices

-   specify ncol, and/or byrow=TRUE to fill by rows
-   Generally it is by columns.
-   Matrix multiplication operator %*%

# 
-   summary(): vector-style summary of column

`apply()`, takes 3 arguments:

- the array or matrix
- then 1 for rows and 2 for columns, and so on.
- then a name of the function to apply to each
